### 2025-07-21 paper/quarto_aaai/paper.tex
Clarify some sentences

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
feature mutability constraints. Much existing research has therefore
focused on developing post-hoc methods to generate counterfactuals that
meet these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals during the training phase to minimize the divergence
between learned representations and plausible, actionable explanations.
We demonstrate empirically and theoretically that our proposed method
facilitates training models that deliver inherently desirable
explanations and have improved adversarial robustness.
\end{abstract}
\section{Introduction}\label{introduction}

@@

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. The authors propose a way
to generate counterfactuals that are maximally faithful models in that
they are consistent with what models have learned about the underlying
data. Formally, they rely on tools from energy-based modelling
\citep{teh2003energy} to minimize the contrastive divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by a model. Their counterfactual explainer, \emph{ECCCo},
yields plausible explanations if and only if the underlying model has
learned representations that align with them. The authors find that both
deep ensembles \citep{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \citep{grathwohl2020your}, a form of constrastive
learning, tend to do well in this regard.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as follows:

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but this is not guaranteed if the assumptions in Prp.
\ref{prp-mtblty} are violated.}

@@

\emph{Plausibility is costly.} As noted elsewhere
\citep{altmeyer2024faithful}, more plausible counterfactuals are
inevitably more costly. Thus, while CT improves plausibility and
robustness, it can impact average costs and validity as cheap,
implausible and adversarial explanations are removed from the solution
space.

\emph{CT increases the training times.} Like contrastive and robust
learning, CT is more resource-intensive than conventional regimes. Three
factors mitigate this effect: (1) CT yields itself to parallel
execution; (2) it amortizes the cost of CEs for the training samples;
and (3) it can be used to fine-tune conventionally-trained models.

@@

Secondly, it is worth extending the analysis done here to additional CE
generators including those that do not rely on computationally costly
gradient-based optimization. This would reduce training costs while
potentially maintaining benefits of CT.

>>>>

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
feature mutability constraints. Much existing research has therefore
focused on developing post-hoc methods to generate counterfactuals that
meet these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals during the training phase to minimize the divergence
between learned representations and plausible, actionable explanations.
We demonstrate empirically and theoretically that our proposed method
facilitates training models that deliver inherently desirable
counterfactual explanations and exhibit greatly improved adversarial
robustness.
\end{abstract}
\section{Introduction}\label{introduction}

@@

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. The authors propose a way
to generate counterfactuals that are maximally faithful in that they are
consistent with what models have learned about the underlying data.
Formally, they rely on tools from energy-based modelling
\citep{teh2003energy} to minimize the contrastive divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by a model. Their algorithm, \emph{ECCCo}, yields
plausible counterfactual explanations if and only if the underlying
model has learned representations that align with them. The authors find
that both deep ensembles \citep{lakshminarayanan2016simple} and joint
energy-based models (JEMs) \citep{grathwohl2020your}, a form of
constrastive learning, tend to do well in this regard.

@@

We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as:

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but this is not guaranteed if the assumptions in Proposition
\ref{prp-mtblty} are violated.}

@@

\emph{Plausibility is costly.} As noted by \citet{altmeyer2024faithful},
more plausible counterfactuals are inevitably more costly. CT improves
plausibility and robustness, but it can impact average costs and
validity when cheap, implausible and adversarial explanations are
removed from the solution space.

\emph{CT increases the training times.} Just like contrastive and robust
learning, CT is more resource-intensive than conventional regimes. Three
factors mitigate this effect: (1) CT yields itself to parallel
execution; (2) it amortizes the cost of CEs for the training samples;
and (3) it can be used to fine-tune conventionally-trained models.

@@

Secondly, our analysis covers CE generators with different
characteristics, but it is interesting to extend it to more algorithms,
including ones that do not rely on computationally costly gradient-based
optimization. This should reduce training costs while possibly
preserving the benefits of CT.

### 2025-07-21 paper/sections/experiments.qmd
Clarify some sentences

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Prp. \ref{prp-mtblty} are violated.*

>>>>

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Proposition \ref{prp-mtblty} are violated.*
